Thermal reactionomes reveal adaptive responses to thermal extremes in warm

### 2014-10-20 doc/ApTranscriptome_manuscript.Rmd
more final tweaks

---
title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we describe a novel analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 8% of all genes were thermally responsive. The majority of upregulation was specific to low temperatures, and consistent with the *enhanced defense hypothesis* the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptation. To better understand the relationship between these two mechanisms, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. Surprisingly, we found divergent transcriptomic responses in these species. *Aphaenogaster picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors.

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Many species, especially small-bodied arthropods, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010]. Thermal defense strategies are shaped by the environmental regime organisms experience [@angilletta_jr2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the evolution of thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes. This information is critical for understanding the adaptive potential of species to future climates [@hoffmann2008].

@@

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* (mean annual temperature (MAT) at collection site = 14.6°C, seasonality = 7,678) experiences significantly warmer and less seasonally variable temperatures than *A. picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A. carolinensis* versus 41.4°C for *A. picea*; Supplemental Material) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*; Supplemental Material) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material).

@@

options(digits = 2)
panderOptions('round', 2)
panderOptions('keep.trailing.zeros', TRUE)
```

@@

pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
             big.mark = ",", split.tables = Inf, 
             caption = "Table of the number of thermally-responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally responsive in one species but not the other.")
```

The total number of thermally-responsive transcripts did not differ between species (...stat...). Within this set, there was significant qualitative and quantitative variation in thermal responses; both among transcripts and between the two species. We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C). The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A. carolinensis* and *A. picea*, respectively), followed by **Intermediate** (29%), **High** (12%),**Bimodal** (10%) and **Not Responsive** (8%) for *A. carolinensis*, and **Bimodal** (17%), **High** (14%), **Intermediate** (10%) and **Not Responsive** (6%) for *A. picea*.

@@

Despite their similar total numbers of responsive transcripts, the two species differed in the proportions of transcripts in each response category (Table 1). *Aphaenogaster picea* upregulated significantly more genes in response to both temperature extremes (*Bimodal*, ...stats...), whereas *A. carolinensis* downregulated more transcripts under these conditions (*Intermediate*, ...stats...). Consistent with the *enhanced defense* hypothesis, *A. picea* was more responsive specifically to low temperatures, inducing expression of 1,146 (30%) more transcripts than *A. carolinensis* (Table 1, ... stats...). In contrast, at high temperatures, *A. carolinensis* matched the predictions of the *tolerance hypothesis*, inducing expression of ~14% fewer transcripts than *A. picea* (Table 1).

@@

Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperature extremes: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig. 2A). Moreover, the transcripts upregulated in *A. picea* included nearly the entire set upregulated in *A. carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Supplemental...). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* might need to induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the short growing season that would necessitate foraging under broader ranges of temperature (...stuble data...).

@@

In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A. carolinensis*. At least 10% of their transcriptomes are thermally responsive. For both species, more transcripts responded to low than high temperatures. Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential in response to high than low temperatures. Our results also indicate that these two ant species have different responses to thermal extremes. *A. picea* responds by increasing metabolism and activating heat shock proteins and other protective molecules, whereas *A. carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolerance. The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species. Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.

@@

We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina (36.037° N -78.874° W). The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice DeMarco, pers. communication), and there is about 2% pairwise divergence between these clades using five nuclear genes (Bernice DeMarco, unpublished data). Colonies of these species were maintained in the lab for 6 months prior to experimentation.

>>>>

---
title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: pnas1.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we describe a novel analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis* in a heat-shock experiment. We found that at least 8% of all genes were thermally responsive. The majority of upregulation was specific to low temperatures, and consistent with the *enhanced defense hypothesis* the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptation. To better understand the relationship between these two mechanisms, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. We found substantial differences in their patterns of gene expression. *Aphaenogaster picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors.

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Many species, especially small-bodied arthropods, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010]. Thermal defense strategies are shaped by the environmental regime organisms experience [@angilletta2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the evolution of thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes. This information is critical for understanding the adaptive potential of species to future climates [@hoffmann2008].

@@

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences warmer (mean annual temperature (MAT) at collection site = 14.6°C) and less variable (temperature seasonality = 7,678) temperatures than *A. picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A. carolinensis* versus 41.4°C for *A. picea*; Supplemental Material) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*; Table S1) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

@@

options(digits = 3)
panderOptions('round', 3)
panderOptions('keep.trailing.zeros', TRUE)
```

@@

pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
             big.mark = ",", split.tables = Inf, 
             caption = "Table of the number of thermally-responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are not thermally responsive in one species but are in the other species.")
```

Though the total number of thermally-responsive transcripts did not differ between species ($\chi^2$~`r chi1$parameter`~ = `r round(chi1$statistic, 2)`, *P* = `r round(chi1$p.value, 2)`) there was significant qualitative and quantitative variation in thermal responses among transcripts and between the two species. To examine this, we grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C). The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A. carolinensis* and *A. picea*, respectively), followed by **Intermediate** (29%), **High** (12%),**Bimodal** (10%) and **Not Responsive** (8%) for *A. carolinensis*, and **Bimodal** (17%), **High** (14%), **Intermediate** (10%) and **Not Responsive** (6%) for *A. picea*.

@@

Despite their similar total numbers of responsive transcripts, the two species differed in the proportions of transcripts in each response category (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 1,259, *P* < 0.001, Table 1, Figure 1). *Aphaenogaster picea* upregulated significantly more genes in response to both temperature extremes (*Bimodal*, post-hoc $\chi^2$ = ...stats...), whereas *A. carolinensis* downregulated more transcripts under these conditions (*Intermediate*, post-hoc $\chi^2$ = ...stats...). Consistent with the *enhanced defense* hypothesis, *A. picea* was more responsive specifically to low temperatures, inducing expression of 1,146 (30%) more transcripts than *A. carolinensis* (Table 1, $\chi^2$ = ...). In contrast, at high temperatures, *A. carolinensis* matched the predictions of the *tolerance hypothesis*, inducing expression of ~14% fewer transcripts than *A. picea* (Table 1).

@@

Under the *enhanced defense* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperature extremes: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, the two species upregulated responsive transcripts at the same threshold temperature (Fig. 2A). Moreover, the transcripts upregulated in *A. picea* included nearly the entire set upregulated in *A. carolinensis* as well as an additional set of ~1,000 transcripts (Table 1), enriched for metabolism and damage repair functions (Supplemental...). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* might need to induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the short growing season that would necessitate foraging under broader ranges of temperature (...stuble data...).

@@

In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A. carolinensis*. At least 10% of their transcriptomes are thermally responsive. For both species, more transcripts responded to low than high temperatures. Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have more limited genetic accomodation to high than low temperatures. Our results also indicate that these two ant species have different responses to thermal extremes. *A. picea* responds by increasing metabolism and activating heat shock proteins and other protective molecules, whereas *A. carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolerance. The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species. Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.

@@

We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina (36.037° N -78.874° W). The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice DeMarco, pers. communication), and there is about 2% pairwise divergence between these clades using five nuclear genes (Bernice DeMarco, unpublished data). Colonies of these species were maintained in common conditions at ?? for 6 months prior to experimentation.
